---
title: "Mini-Project 2"
author: "Vivian Wang, Maggie Wang"
date: "October 31, 2017"
output: html_document
---

# Distribution of Donations to House Candidates

> From this data, we can learn about the distribution of amounts donated by individuals to campaign committees of different party candidates for the House of Representatives. 


## Notes

* The threshold for an individual’s donation to be recorded in the 2011-2012 election cycle was $200. 

* The donations were counted into one of 6 categories based on monetary amount: $200-$500, $501-$1000, $1001-$2000, $2001-$5000,$5001-$10,000, and over $10,000.

```{r setup, include=FALSE, message=FALSE}
library(tidyverse)
library(dplyr)
```

```{r, include=FALSE}
load("house_elections.rda")
load("candidates.rda")
load("committees.rda")
load("contributions.rda")
```

```{r warning=FALSE}
# Cleaning the data
ind_can_prty <- contributions %>% 
  select(cmte_id, transaction_amt) %>% 
  left_join(committees, by = "cmte_id") %>%                   #adding the committees information to donations
  select(cmte_id, transaction_amt, cmte_type, cand_id) %>%
  filter(cand_id!="") %>%                                     #only examine at the campaign committees
  filter(cmte_type == "H") %>%                                #further limit to those affiliated to candidates for the House
  left_join(candidates, by = "cand_id") %>%                   #obtain the party affiliation of candidates
  select(transaction_amt, cmte_id, cand_id, cand_name,
         cand_party_affiliation)

# Dividing the donation amounts into categories
# breaks decided from some exploration of data, process omitted
ind_can_prty$category <- cut(ind_can_prty$transaction_amt, 
                     breaks = c(-Inf, 500,1000,2000,5000,10000, Inf), 
                     labels = c("200-500","501-1000","1001-2000",
                                "2001-5000","5001-10000","Over 10000"))

# Further prepare the data for graphing
ind_can_prty <- ind_can_prty %>% 
  mutate(cand_party_affiliation = factor(cand_party_affiliation), 
         category = factor(category)) %>% 
  select(cand_party_affiliation, category) %>% 
  filter(cand_party_affiliation != "UNK") %>%        #filter out the donation that went to candidates without party info
  group_by(cand_party_affiliation, category) %>% 
  summarize(n = n())
```

```{r}
# Function to create plot
plot_prty <- function(df){
  df <- df  %>% arrange(cand_party_affiliation)
  df %>% ggplot(aes(x = cand_party_affiliation, y = n, fill = category)) + 
    geom_bar(position = "fill", stat = "identity", width = .5) + 
    scale_fill_brewer(name = "Donation in dollars", 
                      palette = "Pastel1") + 
    xlab("Party of Committee's Affiliated Candidate") + 
    ylab("Percentage of Donors") + 
    scale_y_reverse()
}
```


> The bars show the percentage of donations that came from each category for each party affiliations.

## The Democratic Party

```{r}
# Plotting the data for the Democratic Party
dem_ind_can_prty <- ind_can_prty %>% 
  filter(cand_party_affiliation == "DEM")
plot_prty(dem_ind_can_prty) + 
  scale_x_discrete(label = "Democratic")
```

### Here's what we found:

* The greatest proportion of donations to Democratic candidates fell within the $501-$1000 range. The next highest proportion of donations was within the $1001-$2000 range. 

* The smallest proportion of donations fell within the $2001-$5000 range while no individual donations exceeded $5001. Surprisingly, a relatively small proportion of donations were in the smallest-amount category ($200-$500). We expected this category to have a much higher percentage as historically, the economic backgrounds of individuals favoring the Democratic party are lower in comparison to other parties such as the Republicans.  


## What about the other parties?

This graphic displays the distribution of donation amounts to campaign committees of House candidates from 5 party affiliations: Democratic, Republican, Green, Democratic Farmer Labor, and independent. 

```{r}
# Comparing all five parties together
plot_prty(ind_can_prty) + 
  scale_x_discrete(label = c("Democratic", "Dem-Farm-Labor", 
                             "Green", "Independent", 
                             "Republican"))
```


### Here is what we found:

* Compared to the other parties, the Green candidates by far had the most smallest-amount donations ($200-$500).

* The Republican candidates had the most donations with amounts over $2,000; in fact, they had the most individual donations of amounts beyond $10,000. 

* Interestingly, the Democrats and Republicans had similar proportions of small-amount donations ($200-$500 and $501-$1000) which was unexpected because we had thought that the Republicans would have a much smaller proportion of small-amount donations. Instead, the Republicans actually had a higher proportion of $200-$500 donations than the Democrats. 

* Finally, independent candidates had the highest percentage of $1001-$2000 donations amongst the parties. This was interesting because we had hypothesized that candidates from more established parties like the Republican Party would have a greater proportion of donations within this relatively high monetary range.